Significant wind disturbances induced by giant dunes.

### 2021-12-15 Paper/main.tex
Rewriting Introduction

Whenever a flow encounters an obstacle, various types of interactions can arise depending the different time and lengths scales involved. In the case of atmospheric flows, this depends mainly on the part of the vertical structure of the atmosphere, schematically composed of a turbulent boundary layer topped by a turbulence-free part, with which the obstacle interacts \citep{Stull1988}.
%
At the largest scale, the feedback of mountains on the stratified flow of the free atmosphere results in wave generation as well as significant wind disturbances, such as foehn winds in the lee side \citep{refs}. Inside the boundary layer, the interaction between a turbulent flow and hilly surfaces is for example key to the understanding ocean surface wind-driven waves, or eolian bedforms in desert \citep{Belcher1998, Sullivan2010, Courrech2015}.

Indeed, eolian sand dunes typically emerge from the feedback of the topography on the turbulent flow, which speeds up close to the dune crest \citep{Rubin1987, Charru2013, Courrech2014}. Later on, when the dune reaches an intermediate size, it may also induce significant wind deflections. This can impact the sediment pathways of coastal systems \citep{Hesp2015}, or affect the collective behavior of dune populations with long-range interactions due to flow disturbances induced by each individual \citep{Smith2017, Bacik2020}. As the dunes increase in size by collisions and coarsening, they sometimes reach a giant size, comparable to the boundary layer depth, thus inducing interactions not only with the turbulent flow of the ABL, but also with the free atmosphere~\citep{andreotti2009}. However, the wind disturbances induced by these giant dunes have never been quantified.

The study of the impact of obstacles on the atmospheric flows allow its incorporation within meteorological numerical model. Therefore, they mainly become limited by the precision of the included topographical data, as well as the spatial grid of the model. For example, the latest climate reanalysis, ERA5-Land, is limited by its $9~\textup{km}$ spatial resolution, while including the data $30$-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007, munoz2021}. As such, it can not reproduce the flow disturbances induced by giant dunes, which have a typical length scale $\sim 1~\textup{km}$.

Here, we compare the wind predictions from the ERA5-Land dataset to local measurements in four different places across the Namib desert. In places with no significant topographies smaller than the model grid, we show that both wind datasets agree with each other. On the contrary, in places with giant dunes, we show that they may differ for some specific meteorological conditions, that we link to the circadian cycle of the ABL. We thus highlight the importance of the mid-scale topographies for local wind regimes, and its implications in the case of sand seas for smaller-scale eolian bedforms.

>>>>

Whenever a flow encounters an obstacle, different interactions can arise depending on the different time and length scales involved. In the case of atmospheric flows, this mainly depends on the part of the atmosphere, schematically composed of a turbulent boundary layer topped by a part free of turbulence, with which the obstacle interacts \citep{Stull1988}.
%
At the largest scale, the feedback of mountains on the stratified flow of the free atmosphere results in wave generation as well as significant wind disturbances, such as downslope winds in the lee side \citep{Duran1990}. Inside the boundary layer, the interaction between a turbulent flow and hilly surfaces is for example key to the understanding ocean surface wind-driven waves, or eolian bedforms in desert \citep{Belcher1998, Sullivan2010, Courrech2015}.

Indeed, eolian sand dunes typically emerge from the feedback of the topography on the turbulent flow, which speeds up close to the dune crest \citep{Rubin1987, Charru2013, Courrech2014}. Later on, when the dune reaches an intermediate size, it may also induce significant wind deflections. For exemple, this can affect the sediment pathways of coastal systems \citep{Hesp2015}, or the collective behavior of dune populations, through long-range interactions due to flow disturbances induced by each individual \citep{Smith2017, Bacik2020}. As the dunes increase in size by collisions and coarsening, they sometimes reach a giant size comparable to the boundary layer depth~\citep{andreotti2009}. However, the wind disturbances induced by these giant dunes have never been quantified.

The study of the feedback of obstacles on atmospheric flows allows its incorporation into numerical meteorological models. The latter then become mainly limited by the accuracy of the included topographical data, as well as by the spatial grid of the model. For example, the ERA5-Land climate reanalysis is limited by its $9~\textup{km}$ spatial resolution, while including the data $30$-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007, munoz2021}. As such, it can not reproduce the flow disturbances induced by giant dunes, which have a typical length scale $\sim 1~\textup{km}$.

Here, we compare the wind speed and direction of the ERA5-Land dataset to local measurements in four different places across the Namib desert. Where no significant topographies smaller than the model grid are present, we show that the two wind datasets agree with each other. On the contrary, in places with giant dunes, we show that they may differ for some specific meteorological conditions, that we link to the circadian cycle of the atmospheric boundary layer. In doing so, we highlight the importance of medium-scale topographies for local wind regimes and, in the case of sand seas, draw implications for smaller-scale eolian bedforms.
